### 2016-01-20 Manuscript/SW_early_life_telomeres.Rmd
More post evolution edits

**Telomeres reveal silver spoon effects in a wild population**

Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^, and David S. Richardson^1,5,^.

@@

###Abstract
Understanding the links between early-life experiences and later-life survival ("silver spoon effects") is fundamental to understanding life-history evolution. Telomeres, the protective caps at the ends of chromosomes, shorten  in response to oxidative stress, and telomere shortening is correlated with reduced survival. Thus, telomere dynamics may help us quantify individual variation in early-life costs, and thus enhance our understanding of how and why silver spoon effects occur. We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), across multiple cohorts spanning 14 years. We show that telomere lengths and rates of shortening vary across cohorts, and that individuals born in breeding seasons with higher food availability had longer telomeres and reduced telomere shortening. Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which  represents nestling growth), and positively related to the number of helpers present in the natal territory. Finally, we found that increased survival later in life was associated with longer telomeres and reduced telomere shortening rates in early life. Our results show that telomeres can both measure variation in early-life costs and predict late-life survival in wild populations.

@@

Telomere lengths were measured in `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples. Of these, `r nrow(FlSA)+nrow(chicks)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`; `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings). For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken within two years of the original catch.  We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay  [full details in @Barrett2012].

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin Island over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012]. This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics. To overcome this, we analysed how variation in early-life telomere length was related to variation in early-life conditions both among and within breeding seasons.

First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated mean telomere length among all birds born in a given (major or minor) breeding season (i.e. a cohort), excluding cohorts where the sample size $\leq$ 5 birds (leaving a sample size of n = `r nrow(juvseason)` cohorts). We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.

We then conducted individual-level analyses to test how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts, using general linear models. For these analyses, both response and explanatory variables were standardised by mean-centring within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment. As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory. We also included interactions between age class and all the other variables. We used model averaging to assess how these variables shape telomere length and dynamics in early life. We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

For individuals with longitudinal data (*n* = `r nrow(Loss)`) we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013]. Using this dataset we repeated the above analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.

Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts. First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean lifespan of all individuals within each cohort with mean telomere length and rate of loss in that cohort, using linear regression. We then used parametric survival analysis to test whether survival was related to individual-level telomere length and rate of loss. For the latter analysis, again we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival. The survival time distribution was chosen by comparing the fit of the observed survival data against a range of distributions (exponential, Weibull, log-normal, gamma, Gompertz and generalized F) by visually inspecting plots and comparing AIC values. As explanatory variables of survival, we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis. With this approach we aimed to assess the relationship between early-life telomere length and survival, while controlling for potentially confounding variables that might be correlated with telomere dynamics.

@@

#####Early-life telomere length and age
Mean ($\pm$ standard error) telomere length in first-year birds was longer than in adults of any age (Fig. 1A) We also found substantial age-related variation in telomere length within the first year of life, with longer telomeres in nestlings compared to fledglings (ANOVA; F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`; Fig. 1B).

#####Among-cohort variation in early-life telomere dynamics
Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), with a tendency toward shorter telomeres in more recent years (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Tarsus,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig. 2). Variation in mean telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Tarsus,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig. 2A). Running the regression of cohort-level telomere length and food availability separately for the two age classes revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`).

The rate of telomere shortening in early life (taken from the smaller subset of birds for which we had longitudinal samples) also varied significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(LayYear),data = Loss))[1,4])`; P = `r round2(anova(lm(TROC~factor(LayYear),data = Loss))[1,5])`), and rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`). This pattern corresponded with the cross-sectional data: individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (Fig. 2B). Unfortunately we did not have sufficient sample sizes within cohorts to run this analysis separately between age classes.

#####Within-cohort variation in early-life telomere dynamics
The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and the number of helpers present in the natal territory, as well as interactions between age class and both tarsus length and the number of helpers (Table S1). This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=juvall))-AICc(lm(cenTL~Fledged*(cenTarsus+cenHelper),data=juvall)))`; R^2^ = `r round2(summary(lm(cenTL~Ageclass*(cenTarsus+cenHelper),data=juvall))$r.squared)`). Both tarsus length and the number of helpers had high relative importance in the top model set (Fig. 3A). Telomere length was positively related to the number of helpers in nestlings, but not in fledglings (Fig. 3B). Tarsus length was negatively related to telomere length in nestlings, but not in fledglings (Fig. 3C).

For the subset of individuals for which we had longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e. tarsus length and number of helpers) were also associated with differences in rates of early-life telomere loss. Due to a limited sample size, we considered all age classes together for this analysis. The number of helpers in the natal territory was negatively associated to the rate of telomere loss (). We found no relationship between tarsus length and telomere loss ().

@@

library(survival)
m1 <- coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+GroupSize+Fledged+frailty(LayYear)))

A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(juv,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(juv,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including all first-year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB+LayYear), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`). Running this analysis between age classes revealed differential survival effects. Nestling telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`; Fig. 5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3], lessthan = F)`). This effect was positive, with increased survival later in life associated with longer telomeres (Fig. 6B). Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).

@@

Our finding that early-life telomere length and rate of shortening are related to food availability adds to a growing body of evidence from humans and wild animals showing that the natal environment can have pronounced impacts on early-life telomere dynamics [reviewed in @Price2013; @Monaghan2014].  However, very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015], making it impossible to statistically evaluate whether this variation is driven by temporal variation in the environment. The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in environmental conditions does affect telomere dynamics, and that this effect can be seen at the population level. Interestingly, in the Seychelles warbler survival is not directly linked to annual variation in food availability [@Brouwer2006; @Hammers2013]. It has been suggested that telomeres act as biomarkers of cost in natural populations, as they are able to capture individual and cohort-level variability in the costs experienced as a result of environmental pressures [@Asghar2015]. Our results show that telomeres can detect hidden costs in natural populations that are not detectable using life-history and environmental data alone.

@@

We found that, within cohorts, tarsus length and the number of helpers were the most important predictors of telomere length in Seychelles warblers, and that these effects occurred specifically at the nestling stage. The tarsus effect probably reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976]. Telomere loss is most rapid early in life due to ongoing cell replication [@Frenck1998], and a negative correlation between telomere length and body size in early life is therefore expected.

Ours is the first study to show that the presence of helpers has a beneficial effect on offspring in terms of their telomeres. In the Seychelles warbler the survival benefits to a nestling of having helpers in the natal territory is well established [@Komdeur1994;@Brouwer2012], and that the beneficial effect of helpers is detectable in terms of differential telomere length, is therefore encouraging. However, we found no evidence that telomere loss was related to the number of helpers - this is perhaps surprising as telomere loss is expected to be a better indicator of stress than telomere length *per se*, and effects of telomere length may occur as a result of correlated to telomere loss [@Boonekamp2014]. However, our longitudinal dataset was limited, both in terms of sample size and resolution (i.e. time between sampling events). Seychelles warblers are rarely sampled more than once within their first year of life, so much of the telomere shortening that occurs in early life will be missed with our sampling regime. It is likely, therefore that only very strong effects of environmental variation on telomere loss will  be detected in this dataset. Thus telomere length constitutes a better indicator of early-life stress.

While the relationship between mortality and telomeres in adults has been established for some time [@Cawthon2003], only recently has the link between later-life survival and early-life telomere dynamics been studied. In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics. Our results are, to our knowledge, the first to demonstrate that later-life survival is related to early-life telomere length in a wild population. This effect can be seen at the individual level, and (to a lesser extent) at the cohort level, although the nature of the relationship varies according to the age class in which telomeres are measured. At the individual level, we found that telomere length was related to survival in fledglings, but not nestlings. This is not surprising, as only by the fledgling stage is telomere length likely to reflect the accumulation of early-life stress (see also previous paragraph). The relationships between lifespan and population-level telomere length were in the expected direction (Fig. 4), but not significant, which is perhaps not surprising given that even though we have a long-term dataset, we are still restricted in terms of sample size when conducting cohort-level analyses. Nonetheless, our data support the general conclusion that early-life telomere dynamics are important indicators of early-life stress.

There are numerous promising avenues for future research into telomere evolutionary ecology in this system and others. Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014]. Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length predicts lifetime reproductive success. A further avenue for future research will be examining, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014]. We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.

>>>>

**Strong cohort effects on early-life telomere length in a wild population**

Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Martijn Hammers^2^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^, and David S. Richardson^1,5,^.

@@

###Abstract
Understanding the short and long term costs of individual early-life experiences is fundamental to understanding life-history evolution. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival. Thus, telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival. We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), across multiple cohorts spanning 14 years. we found that, in accordance with other studies, telomere length and loss are greatest in early life. However, we also show that juvenile telomere lengths varies markedly among cohorts. We found no evidence that early-life social environment (number of helpers, group size) or ecological conditions (territory quality) were related to telomere length, although we found tentative evidence that telomere length varied among summer and winter breeding season. Finally, we found that increased survival later in life was associated with longer telomeres and reduced telomere shortening rates in early life, but these effects could . Our results highlight the inmportance of cohort effects in studies of telomere length, and suggest that

@@

Telomere lengths were measured in `r nrow(juv)+nrow(Loss)` samples, . Of these, `r nrow(juv)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`. For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken within two years of the original catch.

We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay [@Barrett2012]. Post-qPCR processing of samples was carried out as in Barrett *et al.* [-@Barrett2012], with the following amendments. First, averaging of technical repeats was carried out using custom made R scripts (available as supplementary material). Second, a change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^oC to 61^oC. This resulted in consistently higher, but repeatable, CQ values for the telomere reaction, so we implemented a simple correction to ensure that telomere lengths were repeatable across annealing temperatures (supplementary material). Finally, we excluded telomere lengths greater than 30kb, which maximised repeatability of our own data [see also @Bauch2012]. Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011].

For the subset of samples with longitudinal data we calculated a rate of telomere loss between the first-year and adult samples by subtracting adult telomere length from early-life telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. We used  general linear mixed models along with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts. As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory. As random effects we included birth year and qPCR plate ID. We first created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable. In the analysis of telomere loss we also included initial telomere length as an explanatory variable, as longer telomeres have been shown to decrease in length more rapidly, even after correcting for regression to the mean [@Verhulst2013]. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

We used Cox regression to test whether survival was related to individual-level telomere length and rate of loss. We ran these models with and without lay year as a frailty term to test whether any effects on telomere length on survival could be explained by cohort effects.

@@

#####Early-life telomere length and age
Inter-plate repeatability of telomere length, based on `r length(table(dd0[duplicated(dd0$BloodID),]$BirdID))` samples measured at least twice, was 0.78 (CI = 0.73-0.82). Mean ($\pm$ standard error) telomere length in first-year birds was longer than in adults of any age (Fig. 1A). We also found substantial age-related variation in telomere length within the first year of life, with longer telomeres in nestlings compared to fledglings (ANOVA; F = `r round2(anova(lm(TLKB~Fledged,data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~Fledged,data = juv))[1,5])`; Fig. 1B).

#####Cohort-level variation in early-life telomere dynamics
Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; P = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), with a tendency toward shorter telomeres in more recent years (linear regression of mean telomere length and lay year, R^2^ = `r round2(summary(lm(TLKBmean~LayYear,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TLKBmean~LayYear,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round3(anova(lm(TLKBmean~LayYear,data = juvseason,na.action=na.exclude))[1,5])`; Fig. 2). Variation in mean telomere length over breeding seasons was not related yearly variation in island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(TLKBmean~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(TLKBmean~Insect,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(TLKBmean~Insect,data = juvseason,na.action=na.exclude))[1,5])`).

#####Individual-level variation in early-life telomere dynamics
The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and season (summer vs winter) (Table S1). This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lmer(LogTL~1+(1|FieldPeriodID)+(1|PlateID),data=juv,REML = FALSE))-AICc(lmer(LogTL~Fledged+Tarsus+Season+(1|FieldPeriodID)+(1|PlateID),data=juv,REML = FALSE)))`). All three variables importance in the top model set, but only ageclass had confidence intervals that did not overlap zero (Fig. 3A). Telomere length was slightly longer in juveniles born in winter compared to summmer seasons, and this effect appeared most pronounced in nestlings (although sample size of nestlings born in winter seasons is very small) (Fig. 4A). Tarsus length was negatively related to telomere length, although much of this effect could be explained by ageclass, as fledglings have both longer tarsi and shorter telomeres compared to nestlings  (Fig. 4B).

The longitudinal data showed the same patterns as the cross-sectional data. The top model explaining telomere loss again contained age class, tarsus length and season, although in this instance the top model was only a marginally better fit than the null model ($\Delta$AICc = `r round3(AICc(lmer(TROC~(1|LayYear),data=Loss,REML = FALSE))-AICc(lmer(TROC~Fledged+Tarsus+Season+(1|LayYear),data=Loss, REML = FALSE)),lessthan=F)`). Fledglings lost telomeres at a lower rate compared to nestlings, while birds with longer tarsi and birds born in winter seasons had higher rates of telomere shortening compared to those born in the summer (Fig. 3B). However, it is important to note that sample size for this longitudinal analysis was small and all confidence intervals overlapped zero (Fig. 3B).

@@

A survival model including only early-life telomere length revealed a significant and positive effect on survival (estimate = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL)))$coefficients[1])`, P = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL)))$coefficients[5])`; Fig. 2). However, when we controlled for cohort effects by including lay year as a frailty term in the model, telomere length was no longer a significant predictor of survival (estimate = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+frailty(LayYear))))$coefficients[1,1])`, P = `r round3(summary(coxph(with(juv,Surv(Lifespan+1,Died)~LogTL+frailty(LayYear))))$coefficients[1,6])`). The longitudinal data showed that the amount of telomere shortening experienced in early life had no effect on survival, regardless of whether we included lay year (estimate `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC+frailty(LayYear))))$coefficients[1,1])`, P = `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC+frailty(LayYear))))$coefficients[1,6])`), or excluded it (estimate `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC)))$coefficients[1])`, P = `r round3(summary(coxph(with(Loss,Surv(Lifespan+1,Died)~TROC)))$coefficients[5])`).

@@

The clearest result from our study is that telomere length varies markedly among cohorts. Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and the studies that have done so were limited to just two seasons [@Mizutani2013; @Watson2015]. The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in telomere dynamics does occur at the population level.

@@

Despite a large sample size and reasonable repeatability, we found little evidence that any of the social or environmental vairables we measured affect juvenile telomere length in the Seychelles warbler. This is surprising a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival. Future research should therefore examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2014].

We found tentative evidence that, in addition to age class, tarsus length and season had an effect on juvenile telomere length (Figs 3, 4). The tarsus effect most likely reflects the fact that in passerine birds tarsus length is correlated with age during the nestling stage [@Ricklefs1976]. Telomere loss is most rapid early in life due to ongoing cell replication [@Frenck1998], and a negative correlation between telomere length and body size in early life is therefore expected. Indeed, that age class affects both tarsus length and telomere lengths can be clearly seen in our data (Fig. 4B). The difference in telomere length between birds born in summer and winter seasons is more surprising, especially given the directionality of the effect. We find that birds born in winter seasons having longer telomeres (Fig. 4A), while the opp. One possible explanation for this is that parents in good condition are more likely to breed

While the relationship between mortality and telomeres in adults has been established for some time [@Cawthon2003], only recently has the link between later-life survival and early-life telomere dynamics been studied. In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics.

Finally, our study highlights some of the diffculties asosciated with studying telomere dynamics in natural populations.

Our longitudinal dataset was limited, both in terms of sample size and resolution (i.e. time between sampling events). Seychelles warblers are rarely sampled more than once within their first year of life, so much of the telomere shortening that occurs in early life will be missed with our sampling regime. It is likely, therefore that only very strong effects of environmental variation on telomere loss will  be detected in this dataset. Thus telomere length constitutes a better indicator of early-life stress.

There are numerous avenues for future research into telomere evolutionary ecology in this system and others. Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014]. Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length predicts lifetime reproductive success.  We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.

### 2016-01-20 Manuscript/SW_early_life_telomeres_SOM.Rmd
More post evolution edits

#Run the model
lm1 <- lm(cenTL ~ Fledged*(cenHelper + cenNonHelper + Sex + cenTarsus  + cenTQ),
            data = juvall)

@@

#Sort out row/col names
nams <- colnames(t1)
nams[nams == 'cenTarsus'] <- 'Tarsus'
nams[nams == 'cenNonHelper'] <- 'Non helpers'
nams[nams == 'SexMales'] <- 'Sex'
nams[nams == 'cenTQ'] <- 'TQ'
nams[nams == 'cenHelper'] <- 'Helpers'
nams[nams == 'cenHelper:Fledged'] <- 'Ageclass x Helpers'
nams[nams == 'cenNonHelper:Fledged'] <- 'Ageclass x Non helpers'
nams[nams == 'cenTarsus:Fledged'] <- 'Ageclass x Tarsus'
nams[nams == 'cenTQ:Fledged'] <- 'Ageclass x TQ'
nams[nams == 'Fledged:SexMales'] <- 'Ageclass x Sex'
nams[nams == 'Fledged'] <- 'Ageclass'

@@

>>>>

#Run the model
lm1 <- lmer(LogTL ~ Fledged + Tarsus + Helper + NonHelper + Sex + Tarsus + BodyMass  + cenTQ  + Season + (1|LayYear),
            data = juv,
            REML = FALSE)

@@

#Sort out row/col names
nams <- colnames(t1)
#Sort out row/col names
nams[nams == 'NonHelper'] <- 'Non helpers'
nams[nams == 'SexMales'] <- 'Sex'
nams[nams == 'cenTQ'] <- 'TQ'
nams[nams == 'Helper'] <- 'Helpers'
nams[nams == 'BodyMass'] <- 'Body mass'
nams[nams == 'FledgedFledglings'] <- 'Age class'
nams[nams == 'SeasonWinter'] <- 'Season'

@@

```{r Table S2}

#Run the model
lm2 <- lmer(TROC ~ Fledged + Tarsus + Helper + NonHelper + Sex + Tarsus + BodyMass  + cenTQ  + Season + LogTL + (1|LayYear),
            data = Loss,
            REML = FALSE)

#Model averaging

lm3 <- dredge(lm2)
lm3 <- subset(lm3,delta <=6)

#Model averaging
t1 <- as.matrix(subset(lm3, delta<=6))

#Sort out row/col names
nams <- colnames(t1)

nams[nams == 'NonHelper'] <- 'Non helpers'
nams[nams == 'SexMales'] <- 'Sex'
nams[nams == 'cenTQ'] <- 'TQ'
nams[nams == 'Helper'] <- 'Helpers'
nams[nams == 'BodyMass'] <- 'Body mass'
nams[nams == 'FledgedFledglings'] <- 'Age class'
nams[nams == 'SeasonWinter'] <- 'Season'
nams[nams == 'LogTL'] <- 'Initial TL'
